Scaling rates of true polar wander in convecting planets and moons

### 2017-09-11 tpw_rate.tex
Updates for submitting revisions.
Make simple fixes.

\begin{abstract}
Mass redistribution in the convecting mantle of a planet causes perturbations in its moment of inertia tensor. 
Conservation of angular momentum dictates that these perturbations change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
Although the existence of TPW on Earth is firmly established, its rate and magnitude over geologic time scales remain controversial. 
Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
For simple rotating convection, we identify a set of dimensionless parameters that fully characterize true polar wander. We use these parameters to define  timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and relates them to a  broader class of events that enable large and often rapid TPW. We expect these events to have been more frequent in Earth’s past.
\end{abstract}

@@

%%\linenumbers

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
Both of these contributions may also be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
and $C_{ij}$ is the moment of inertia due solely to the internal load.
When the timescale of the surface response is short compared to the true polar wander timescale, it is reasonable to  use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

\citet{tsai2007theoretical} suggested that large TPW events do not always accompany an interchange of the principal moments because the rate of TPW is low. A large 90$^{\circ}$ change in the orientation of the rotation axis can only occur over a long time.  While
this is true for idealized cases, the numerical convection calculation reveals a more complex behavior. 
First, the principal moments can interchange, or almost interchange, quite rapidly. In these cases  the perturbations in the convective moments can become large soon after an interchange. Second, the
 the location of the principal axes do not stay put during a IITPW event. A persistent drift of the principal axes can can lead to extended polar wandering, yielding angular changes greater than 90$^\circ$.

>>>>

\begin{abstract}
Mass redistribution in the convecting mantle of a planet causes perturbations in its moment of inertia tensor. 
Conservation of angular momentum dictates that these perturbations change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
Although the existence of TPW on Earth is firmly established, its rate and magnitude over geologic time scales remain controversial. 
Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
For simple rotating convection, we identify a set of dimensionless parameters that fully characterize true polar wander. We use these parameters to define  timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and relates them to a  broader class of events that enable large and often rapid TPW. We expect these events to have been more frequent in Earth's past.
\end{abstract}

@@

\linenumbers

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
Both of these contributions may also be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
and $C_{ij}$ is the moment of inertia due solely to the internal load.
When the timescale of the surface response is short compared to the true polar wander timescale, it is reasonable to  use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

\citet{tsai2007theoretical} suggested that large TPW events do not always accompany an interchange of the principal moments because the rate of TPW is low. A large 90$^{\circ}$ change in the orientation of the rotation axis can only occur over a long time.  While
this is true for idealized cases, the numerical convection calculation reveals a more complex behavior. 
First, the principal moments can interchange, or almost interchange, quite rapidly. In these cases  the perturbations in the convective moments can become large soon after an interchange.
Second, the location of the principal axes do not stay put during a IITPW event. A persistent drift of the principal axes can can lead to extended polar wandering, yielding angular changes greater than 90$^\circ$.
